---
title: "Spatio-temporal Modelling, Lab 9, YYYY-MM-DD"
editor: visual
---

**This exercise will be discussed on YYY-MM-DD.**

# Cokriging

## Prepare the data

::: panel-tabset

## sp

```{r,warning=F}
library(gstat)
library(sp)
library(rgdal)
sp.theme(TRUE)
meuse <- readOGR("./data/meuse.gpkg")
meuse.grid <- as(readGDAL("./data/meuse_grid.tif"),"SpatialPixelsDataFrame")
names(meuse.grid) <- c("part.a","part.b","dist","soil","ffreq")
meuse.grid$soil <- as.factor(meuse.grid$soil)
meuse.grid$ffreq <- as.factor(meuse.grid$ffreq)
g <- gstat(NULL, "logCd", log(cadmium) ~ 1, meuse)
g <- gstat(g, "logCu", log(copper) ~ 1, meuse)
g <- gstat(g, "logPb", log(lead) ~ 1, meuse)
g <- gstat(g, "logZn", log(zinc) ~ 1, meuse)
g
vm <- variogram(g)
vm.fit <- fit.lmc(vm, g, vgm(1, "Sph", 800, 1))
vm.fit
plot(vm, vm.fit)
cor(log(as.data.frame(meuse)[c("cadmium", "copper", "lead", "zinc")]))

```

## sf/stars

```{r}
library(gstat)
library(sf)
library(stars)
library(tmap)
meuse.sf <- st_read("./data/meuse.gpkg")
meuse.grid.stars <- read_stars("./data/meuse_grid.tif")
meuse.grid.stars <- split(meuse.grid.stars, "band")
names(meuse.grid.stars) <- c("part.a","part.b","dist","soil","ffreq")
meuse.grid.stars$soil <- as.factor(meuse.grid.stars$soil)
meuse.grid.stars$ffreq <- as.factor(meuse.grid.stars$ffreq)
bpy.palette=bpy.colors()
g.sf <- gstat(NULL, "logCd", log(cadmium) ~ 1, meuse.sf)
g.sf <- gstat(g.sf, "logCu", log(copper) ~ 1, meuse.sf)
g.sf <- gstat(g.sf, "logPb", log(lead) ~ 1, meuse.sf)
g.sf <- gstat(g.sf, "logZn", log(zinc) ~ 1, meuse.sf)
g.sf
vm.sf <- variogram(g.sf)
vm.fit.sf <- fit.lmc(vm.sf, g.sf, vgm(1, "Sph", 800, 1))
vm.fit.sf
plot(vm.sf, vm.fit.sf)
cor(log(as.data.frame(meuse.sf)[c("cadmium", "copper", "lead", "zinc")]))
```

:::

## Do a cokriging:

::: panel-tabset

## sp

```{r}
cokr = predict(vm.fit, meuse.grid)
summary(cokr)
```

## sf/stars

```{r}
cokr.stars = predict(vm.fit.sf, meuse.grid.stars)
summary(cokr.stars)
```

:::



### *Exercise 9.1* {.unnumbered}
::: {.callout-caution}
## Task

*Explain in language (without equations) what the variables in the resulting object represent, and how they were computed!*
:::


### *Exercise 9.2* {.unnumbered}
::: {.callout-caution}
## Task

*Explain why all the prediction error covariances are positive!*
:::


We want to compute an index of toxicity of the soil, where the different heavy metals are weighted according to their toxicity. The index is computed as

$$I=100\log{Cd}+10logPb+logZn$$

where logCd, logPb and logZn are the log-predictions as obtained above.



### *Exercise 9.3* {.unnumbered}
::: {.callout-caution}
## Task

*Give the R commands to compute this index and to show a map of it. Also give the commands to compute the standard error of the index, taking the error covariances into account.*
:::


### *Exercise 9.4* {.unnumbered}
::: {.callout-caution}
## Task

*How much does the standard error of this toxicity index change as a result of taking the covariances into account, as opposed to ignoring them?*
:::


## Cokriging: the undersampled case

Take a subsample of zinc, called meuse1:

::: panel-tabset

## sp

```{r}
set.seed(1331)
sam = sample(155)
sam1 = sam[1:50]
sam2 = sam[-(1:50)]
meuse1 = meuse[sam1, ]
meuse2 = meuse[sam2, ]
plot(meuse, col = 1, pch = 3)
plot(meuse1, col = 3, pch = 1, add = TRUE)
```

## sf/stars

```{r}
set.seed(1331)
sam = sample(155)
sam1 = sam[1:50]
sam2 = sam[-(1:50)]
meuse1.sf = meuse.sf[sam1, ]
meuse2.sf = meuse.sf[sam2, ]
plot(meuse.sf["zinc"], col = 1, pch = 3,reset=FALSE)
plot(meuse1.sf["zinc"], col = 3, pch = 1, add = TRUE)
```

:::

create a multivariate object with subsample of zinc, and full sample of lead:

::: panel-tabset

## sp

```{r,warning=F}
g = gstat(NULL, "logZinc", log(zinc) ~ 1, meuse1, set = list(nocheck = 1))
g = gstat(g, "logLead", log(lead) ~ 1, meuse, set = list(nocheck = 1))
v12 = variogram(g)
v12 = subset(v12, dist != 0)  # remove 0 distances for fitting
v12.fit = fit.lmc(v12, g, vgm(1, "Sph", 900, 1))
v12.fit
class(v12.fit)
plot(v12, v12.fit)
```

## sf/stars

```{r}
g.sf = gstat(NULL, "logZinc", log(zinc) ~ 1, meuse1.sf, set = list(nocheck = 1))
g.sf = gstat(g.sf, "logLead", log(lead) ~ 1, meuse.sf, set = list(nocheck = 1))
v12.sf = variogram(g.sf)
v12.sf = subset(v12.sf, dist != 0)  # remove 0 distances for fitting
v12.fit.sf = fit.lmc(v12.sf, g.sf, vgm(1, "Sph", 900, 1))
v12.fit.sf
class(v12.fit.sf)
plot(v12.sf, v12.fit.sf)
```

:::

cokrige:

::: panel-tabset

## sp

```{r, warning=F,results=F}
v12.pr = predict(v12.fit, meuse.grid)
spplot(v12.pr, c("logZinc.pred", "logLead.pred"))
spplot(v12.pr, c("logZinc.var", "logLead.var", "cov.logZinc.logLead"))
```

## sf/stars

```{r,warning=F,results=F}
v12.pr.stars = predict(v12.fit.sf, meuse.grid.stars)
tm_shape(v12.pr.stars) + 
  tm_raster(c("logZinc.pred", "logLead.pred"), n=20, title = "", palette = bpy.palette) + 
  tm_facets(free.scales=FALSE) + 
  tm_layout(legend.outside.size = 0.15, panel.labels=c("logZinc.pred", "logLead.pred"))

tm_shape(v12.pr.stars) + 
  tm_raster(c("logZinc.var", "logLead.var", "cov.logZinc.logLead"), n=20, title = "", palette = bpy.palette) + 
  tm_facets(free.scales=FALSE) + 
  tm_layout(legend.outside.size = 0.15, panel.labels=c("logZinc.var", "logLead.var", "cov.logZinc.logLead"))


```

:::

compare the differences of kriging the 50 observations of zinc with the cokriging from 50 zinc and 155 lead observations:

::: panel-tabset

## sp

```{r,warning=F}
v12.fit[1]
v1.pr = predict(v12.fit[1], meuse.grid)
lt = list(list("sp.points", meuse2, pch = 3), list("sp.points", meuse1, pch = 2))
v12.pr$diffpr = v12.pr$logZinc.pred - v1.pr$logZinc.pred
spplot(v12.pr["diffpr"], sp.layout = lt)
v12.pr$diffse = sqrt(v12.pr$logZinc.var) - sqrt(v1.pr$logZinc.var)
spplot(v12.pr["diffse"], sp.layout = lt)
```

## sf/stars

```{r}
v12.fit.sf[1]
v1.pr.stars = predict(v12.fit.sf[1], meuse.grid.stars)
v12.pr.stars$diffpr = v12.pr.stars$logZinc.pred - v1.pr.stars$logZinc.pred
tm_shape(v12.pr.stars) + 
  tm_raster("diffpr", title = "",palette = bpy.palette, n=20, midpoint = NA) + 
  tm_shape(meuse2.sf) + tm_symbols(size = 0.2,  shape = 3, col = "cyan4") + 
  tm_shape(meuse1.sf) + tm_symbols(size = 0.2, shape = 2, col = "cyan4") + 
  tm_layout(legend.outside = TRUE)
v12.pr.stars$diffse = sqrt(v12.pr.stars$logZinc.var) - sqrt(v1.pr.stars$logZinc.var)
tm_shape(v12.pr.stars) + 
  tm_raster("diffse", title = "",palette = bpy.palette, n=20, midpoint = NA) + 
  tm_shape(meuse2.sf) + tm_symbols(size = 0.2,  shape = 3, col = "cyan4") + 
  tm_shape(meuse1.sf) + tm_symbols(size = 0.2, shape = 2, col = "cyan4") + 
  tm_layout(legend.outside = TRUE)

```

:::



### *Exercise 9.5* {.unnumbered}
::: {.callout-caution}
## Task

*Why are the standard errors smaller in the cokriging case?*
:::


## Kriging errors and confidence intervals

Given that we are not certain about zinc levels at locations where we did not measure it (i.e., interpolated grid cells) we can classify them according to their approximate confidence intervals, formed by using mean and standard error (here: on the log scale).

::: panel-tabset

## sp

```{r,warning=F}
# Ordinary Kriging log(zinc)
m <- vgm(0.59, "Sph", 874, 0.04)
# ordinary kriging:
lzn.ok <- krige(log(zinc) ~ 1, meuse, meuse.grid, model = m)
ci = function(cutoff, mean, var, alpha = 0.05) {
    p = 1 - pnorm(cutoff, mean, sqrt(var))  #$
    f = ifelse(p < alpha/2, "lower", ifelse(p < (1 - alpha/2), "not dist.", 
        "higher"))
    factor(f, levels = c("lower", "not dist.", "higher"))
}
lzn.ok$ci250 = ci(log(250), lzn.ok$var1.pred, lzn.ok$var1.var)
lzn.ok$ci500 = ci(log(500), lzn.ok$var1.pred, lzn.ok$var1.var)
lzn.ok$ci750 = ci(log(750), lzn.ok$var1.pred, lzn.ok$var1.var)
lzn.ok$ci1000 = ci(log(1000), lzn.ok$var1.pred, lzn.ok$var1.var)
spplot(lzn.ok, c("ci250", "ci500", "ci750", "ci1000"), col.regions = c("green", 
    "grey", "red"))
```

## sf/stars

```{r}
# Ordinary Kriging log(zinc)
m <- vgm(0.59, "Sph", 874, 0.04)
# ordinary kriging:
lzn.ok.stars <- krige(log(zinc) ~ 1, meuse.sf, meuse.grid.stars, model = m)
ci = function(cutoff, mean, var, alpha = 0.05) {
    p = 1 - pnorm(cutoff, mean, sqrt(var))  #$
    f = ifelse(p < alpha/2, "lower", ifelse(p < (1 - alpha/2), "not dist.", 
        "higher"))
    factor(f, levels = c("lower", "not dist.", "higher"))
}
lzn.ok.stars$ci250 = ci(log(250), lzn.ok.stars$var1.pred, lzn.ok.stars$var1.var)
lzn.ok.stars$ci500 = ci(log(500), lzn.ok.stars$var1.pred, lzn.ok.stars$var1.var)
lzn.ok.stars$ci750 = ci(log(750), lzn.ok.stars$var1.pred, lzn.ok.stars$var1.var)
lzn.ok.stars$ci1000 = ci(log(1000), lzn.ok.stars$var1.pred, lzn.ok.stars$var1.var)
tm_shape(lzn.ok.stars) + 
  tm_raster(c("ci250", "ci500", "ci750", "ci1000"), n=3, title = "", palette = c("green", 
    "grey", "red") ) + 
  tm_facets(free.scales=FALSE, nrow = 1) + 
  tm_layout(legend.outside.size = 0.15, panel.labels=c("ci250", "ci500", "ci750", "ci1000"))
```

:::



### *Exercise 9.6* {.unnumbered}
::: {.callout-caution}
## Task

*For which fraction of the total area is zinc concentration lower than 500 ppm, with 95 percent confidence? Give the number of grid cells for which this is the case, and describe where in the area they are situated.*
:::


For a single threshold, we can vary the confidence level by specifying alpha:

::: panel-tabset
## sp

```{r,warning=F}
lzn.ok$ci001 = ci(log(500), lzn.ok$var1.pred, lzn.ok$var1.var, 0.01)
lzn.ok$ci005 = ci(log(500), lzn.ok$var1.pred, lzn.ok$var1.var, 0.05)
lzn.ok$ci010 = ci(log(500), lzn.ok$var1.pred, lzn.ok$var1.var, 0.1)
lzn.ok$ci050 = ci(log(500), lzn.ok$var1.pred, lzn.ok$var1.var, 0.5)
spplot(lzn.ok, c("ci001", "ci005", "ci010", "ci050"), col.regions = c("green", 
    "grey", "red"))
```

## sf/stars

```{r}
lzn.ok.stars$ci001 = ci(log(500), lzn.ok.stars$var1.pred, lzn.ok.stars$var1.var, 0.01)
lzn.ok.stars$ci005 = ci(log(500), lzn.ok.stars$var1.pred, lzn.ok.stars$var1.var, 0.05)
lzn.ok.stars$ci010 = ci(log(500), lzn.ok.stars$var1.pred, lzn.ok.stars$var1.var, 0.1)
lzn.ok.stars$ci050 = ci(log(500), lzn.ok.stars$var1.pred, lzn.ok.stars$var1.var, 0.5)
tm_shape(lzn.ok.stars) + 
  tm_raster(c("ci001", "ci005", "ci010", "ci050"), n=3, title = "", palette = c("green", 
    "grey", "red") ) + 
  tm_facets(free.scales=FALSE, nrow = 1) + 
  tm_layout(legend.outside.size = 0.15, panel.labels=c("ci001", "ci005", "ci010", "ci050"))
```

:::



### *Exercise 9.7* {.unnumbered}
::: {.callout-caution}
## Task

*Which of these four maps has the highest confidence level, and what is this level? Why are there gray areas on map ci050 at all?*
:::


Now consider the object zn, created under the section of block kriging.

*Hint: If you don't have the znobject in your workspace, you need to reinstantiate it, by executing the complete code chunk, where the object zn has been created!*

::: panel-tabset
## sp

```{r,warning=F, echo=F,results='hide'}
v = variogram(log(zinc) ~ dist, meuse)
zn.res.m = fit.variogram(v, vgm(1, "Sph", 700, 1))
zn = krige(log(zinc) ~ sqrt(dist), meuse, meuse.grid, zn.res.m)
# Ordinary Kriging log(zinc)
m <- vgm(0.59, "Sph", 874, 0.04)
# ordinary kriging:
lzn.ok <- krige(log(zinc) ~ 1, meuse, meuse.grid, model = m)

ci = function(cutoff, mean, var, alpha = 0.05) {
    p = 1 - pnorm(cutoff, mean, sqrt(var))  #$
    f = ifelse(p < alpha/2, "lower", ifelse(p < (1 - alpha/2), "not dist.", 
        "higher"))
    factor(f, levels = c("lower", "not dist.", "higher"))
}

zn = krige(log(zinc) ~ sqrt(dist), meuse, meuse.grid, zn.res.m)
zn$pointkr = zn$var1.pred
zn40 = krige(log(zinc) ~ sqrt(dist), meuse, meuse.grid, zn.res.m, block = c(40, 
    40))
zn400 = krige(log(zinc) ~ sqrt(dist), meuse, meuse.grid, zn.res.m, block = c(400, 
    400))
zn$block40 = zn40$var1.pred
zn$block400 = zn400$var1.pred
zn$pointkr.se = sqrt(zn$var1.var)
zn$block40.se = sqrt(zn40$var1.var)
zn$block400.se = sqrt(zn400$var1.var)

zn$ok.point = ci(log(500), lzn.ok$var1.pred, lzn.ok$var1.var, 0.05)
zn$uk.point = ci(log(500), zn$pointkr, zn$pointkr.se^2, 0.05)
zn$uk.block40 = ci(log(500), zn$block40, zn$block40.se^2, 0.05)
zn$uk.block400 = ci(log(500), zn$block400, zn$block400.se^2, 0.05)
spplot(zn, c("ok.point", "uk.point", "uk.block40", "uk.block400"), col.regions = c("green", 
    "grey", "red"))
```

## sf/stars

```{r}
v.sf = variogram(log(zinc) ~ dist, meuse.sf)
zn.res.m.sf = fit.variogram(v.sf, vgm(1, "Sph", 700, 1))
zn.stars= krige(log(zinc) ~ sqrt(dist), meuse.sf, meuse.grid.stars, zn.res.m.sf)
# Ordinary Kriging log(zinc)
m <- vgm(0.59, "Sph", 874, 0.04)
# ordinary kriging:
lzn.ok.stars <- krige(log(zinc) ~ 1, meuse.sf, meuse.grid.stars, model = m)

ci = function(cutoff, mean, var, alpha = 0.05) {
    p = 1 - pnorm(cutoff, mean, sqrt(var))  #$
    f = ifelse(p < alpha/2, "lower", ifelse(p < (1 - alpha/2), "not dist.", 
        "higher"))
    factor(f, levels = c("lower", "not dist.", "higher"))
}

zn.stars = krige(log(zinc) ~ sqrt(dist), meuse.sf, meuse.grid.stars, zn.res.m.sf)
zn.stars$pointkr = zn.stars$var1.pred
zn40.stars = krige(log(zinc) ~ sqrt(dist), meuse.sf, meuse.grid.stars, zn.res.m.sf, block = c(40, 
    40))
zn400.stars = krige(log(zinc) ~ sqrt(dist), meuse.sf, meuse.grid.stars, zn.res.m.sf, block = c(400, 
    400))
zn.stars$block40 = zn40.stars$var1.pred
zn.stars$block400 = zn400.stars$var1.pred
zn.stars$pointkr.se = sqrt(zn.stars$var1.var)
zn.stars$block40.se = sqrt(zn40.stars$var1.var)
zn.stars$block400.se = sqrt(zn400.stars$var1.var)

zn.stars$ok.point = ci(log(500), lzn.ok.stars$var1.pred, lzn.ok.stars$var1.var, 0.05)
zn.stars$uk.point = ci(log(500), zn.stars$pointkr, zn.stars$pointkr.se^2, 0.05)
zn.stars$uk.block40 = ci(log(500), zn.stars$block40, zn.stars$block40.se^2, 0.05)
zn.stars$uk.block400 = ci(log(500), zn.stars$block400, zn.stars$block400.se^2, 0.05)

tm_shape(zn.stars) + 
  tm_raster(c("ok.point", "uk.point", "uk.block40", "uk.block400"), n=3, title = "", palette = c("green", 
    "grey", "red") ) + 
  tm_facets(free.scales=FALSE, nrow = 1) + 
  tm_layout(legend.outside.size = 0.15, panel.labels=c("ok.point", "uk.point", "uk.block40", "uk.block400"))
```

:::



### *Exercise 9.8* {.unnumbered}
::: {.callout-caution}
## Task

*Compare the fraction classified as lower and higher between ok.point and uk.point, and try to explain the difference!*
:::


### *Exercise 9.9* {.unnumbered}
::: {.callout-caution}
## Task

*Compare the fraction classified as lower and higher between uk.point uk.block40 and uk.block400, and try to explain the difference!*
:::