---
title: "Spatio-temporal Modelling, Lab 2, 2022-10-19"
format: html
editor: visual
---

**This excercise will be discussed on 2022-10-26.**

This exercise is about exploratory time series analysis, fitting periodic components, and AR models. You will need to download the R data file `meteo.Rdata` from the Learnweb and put it into your workspace directory.

## Exploratory Time Series Analysis

w You can load the **RData** file in R using the **load** function. The **class** function allows you to retrieve the class of an R object. In this case, the data is loaded as a [data.frame](http://cran.r-project.org/doc/manuals/r-release/R-intro.html#Data-frames).

```{r}
load("./data/meteo.RData")
class(meteo)
summary(meteo)
```

::: {.callout-note}
## Hint

In case the RData file is not in working directory, you may have to specify the absolute path (e.g. C:/myDirectory/meteo.RData in Windows), but we recommend you to set the working directory properly using the **setwd** function or an R Studio project.
:::

Again, we can generate a plot using the **plot** function:

```{r fig.width=5, fig.height=4}
plot(T.outside ~ date, meteo, type = "l")
```

**T.outside** and **date** are both contained in the meteo data as columns (see the output of **summary** above). In order to state that the temperature outside should be plotted against the dates, we pass the **T.outside \~ date** expression to the plot function as first parameter, then indicate that both are part of the meteo dataset (second parameter), and the third paramter **type="l"** is the same as described above.

## Fitting a periodic component

For fitting the periodic component for Exercise 1, a new function with name **f** is defined. In general, we can define functions in R using the following syntax:

```{r}
funcMultiply = function(x,y){x*y}
z = funcMultiply(3,4)
z
```

In the example above, we define a new function called **funcMultiply** that multiplies two numbers. The two numbers are passed to the function, indicated by **function(x,y)**. If we then invoke funcMulitply with the numbers 3 and 4, we get the result 12.

We may also pass vectors to a function. Taking the example of a multiply function, we may also pass a vector with two numbers.

```{r}
funcMultiplyVector = function(x){x[1]*x[2]}
inputVector = c(3,4)
z = funcMultiplyVector(inputVector)
z
```

*Hint:* In the example for exercises 2.1 and 2.2, the second approach, where a vector is passed to the function is implemented.

The meteo dataset is loaded as a [data.frame](http://cran.r-project.org/doc/manuals/r-release/R-intro.html#Data-frames) in R. The **summary** function lists all columns with summary statistics. We can easily add a new column computed from other using the following syntax:

```{r}
meteo$T.dummy = meteo$T.outside-10
```

The code above creates a new temperature dummy variable called **T.dummy** that consists of the **T.outside** values minus 10. If we now invoke the **summary** function again, the column is listed.

```{r}
summary(meteo)
```

We can now fit a periodic component to the temperature data, using a non-linear optimization **nlm**.

```{r}
#generate the periodic model function and, based on this, 
#the objective function as the sum of squared differences
f = function(x) sum((meteo$T.outside - (x[1]+x[2]*sin(pi*(meteo$hours+x[3])/12)))^2)
#optimize the parameters of the model by using the nlm function
nlm(f,c(0,0,0))
```

#### *Exercise 1* {.unnumbered}
::: {.callout-caution}
## Exercise 1

*How many parameters were fitted?*
:::

We will now plot observations and fitted model together:

```{r fig.width=5, fig.height=4}
#plot the temperature curve first again
plot(T.outside ~ date, meteo, type = "l")

#create a new column in the dataset using the optimized parameters to calculate the new data
meteo$T.per = 18.2 - 4.9 * sin(pi*(meteo$hours+1.6)/12)

#create a new column in the dataset using the optimized parameters to calculate the new data
lines(T.per~date,meteo,col='red')
```

#### *Exercise 2* {.unnumbered}
::: {.callout-caution}
## Exercise 2

*What is the interpretation of the fitted parameters? (if you need to guess, modify them and replot)*
:::

We can now also plot the residual (difference between predicted values and measured temperature values) from this fitted model:

```{r fig.width=5, fig.height=4}
plot(T.outside-T.per~date, meteo, type='l')
title("difference from fitted sinus")
```

Note that a new column **T.per** is generated that contains the predicted values from the model. and that the **lines** function plots the predicted values (shown as red line) in the time series plot generated before.

## Fitting AR models to residuals

Note that the AR models in these exercises are not fitted to the actual outside temperature data (**T.outside**), but to the residuals between the predictions made by the periodic model in exercises 2.1 and 2.2 (**T.per**) and the actual outside temperature values (**T.outside**). Therefore, the residuals are computed by

```{r}
an = meteo$T.outside - meteo$T.per
```

Please use the help functionality of R (and of R Studio) to get more information about the functions **arima**, **acf**, and **tsdiag**. The theoretical background of the two functions is provided in the [slides of our lecture](https://edzer.github.io/mstp/).

The $AR(p)$ model is defined as

$$y_t = \sum_{j=1}^{p} \phi_j y_{t-j} + e_t$$

with $e_t$ a white noise process. For $p=1$ this simplifies to

$$y_t = \phi_1 y_{t-1}+ e_t.$$

Now try to model the residual process as an AR(5) process, and look at the partial correlations.

```{r}
an.ar5 = arima(an, c(5, 0, 0))
an.ar5
```

#### *Exercise 3* {.unnumbered}
::: {.callout-caution}
## Exercise 3

```{r}
acf(an, type = "partial")
```

*Does the **an** process exhibit temporal correlation for lags larger than 0?)*
:::

#### *Exercise 4* {.unnumbered}
::: {.callout-caution}
## Exercise 4

```{r}
acf(residuals(an.ar5), type = "partial")
```

*Does the **residuals(an.ar5)** process still exhibit temporal correlation for lags larger than 0?*
:::

#### *Exercise 5* {.unnumbered}
::: {.callout-caution}
## Exercise 5

*What is the class of the object returned by **arima**?*
:::

#### *Exercise 6* {.unnumbered}
::: {.callout-caution}
## Exercise 6

Let us see what we can do with such an object.

```{r}
methods(class="Arima")
tsdiag(an.ar5)
```

The **methods** function returns the methods of a class, in this case the **Arima** class. The tsdiag function provides time series diagnostic plots.

*Explain what you see in the first two plots!*
:::

## Model selection with Akaike's Information Criterion (AIC)

We can use the **\$aic** suffix to directly retrieve the AIC from an AR model.

```{r}
temp = meteo$T.outside
```

    arima(temp, c(1, 0, 0))$aic
    arima(temp, c(2, 0, 0))$aic
    arima(temp, c(3, 0, 0))$aic
    arima(temp, c(4, 0, 0))$aic
    arima(temp, c(5, 0, 0))$aic
    arima(temp, c(6, 0, 0))$aic
    arima(temp, c(7, 0, 0))$aic
    arima(temp, c(8, 0, 0))$aic
    arima(temp, c(9, 0, 0))$aic
    arima(temp, c(10, 0, 0))$aic

#### *Exercise 7* {.unnumbered}
::: {.callout-caution}
## Exercise 7

*Which model has the smallest AIC?*
:::

#### *Exercise 8* {.unnumbered}
::: {.callout-caution}
## Exercise 8

*Do a similar analysis for the humidity variable in the meteo data set:*

1.  *Fit a periodic trend; give the trend equation.*
2.  *Plot the humidity data and the fitted model.*
3.  *detrend the humidity data to obtain residuals and report for which value of n in an AR(n) model of the anomalies (residuals) the model has the lowest AIC.*
4.  *Up to which lag does the residual humidity process exhibit temporal correlation?*
:::

## Prediction with an AR model

Let us now work with the AR(6) model for the temperature, ignoring the periodic (diurnal) component. Make sure you have "plot recording" on (activate the plot window to get this option).

```{r fig.width=5, fig.height=4}
x = arima(temp,c(6,0,0))
# 10 mins:
plot(meteo$T.outside,xlim=c(9860,9900), type='l')
x.pr = as.numeric(predict(x, 10)$pred)
x.se = as.numeric(predict(x, 10)$se)
lines(9891:9900, x.pr, col='red')
lines(9891:9900, x.pr+2*x.se, col='green')
lines(9891:9900, x.pr-2*x.se, col='green')
title("predicting 10 mins")
# 110 mins:
plot(meteo$T.outside,xlim=c(9400,10000), type='l')
x.pr = as.numeric(predict(x, 110)$pred)
x.se = as.numeric(predict(x, 110)$se)
lines(9891:10000, x.pr, col='red')
lines(9891:10000, x.pr+2*x.se, col='green')
lines(9891:10000, x.pr-2*x.se, col='green')
title("predicting 110 mins")
# 1440 mins, 1 day:
plot(meteo$T.outside,xlim=c(8000,11330), type='l')
x.pr = as.numeric(predict(x, 1440)$pred)
x.se = as.numeric(predict(x, 1440)$se)
lines(9891:11330, x.pr, col='red')
lines(9891:11330, x.pr+2*x.se, col='green')
lines(9891:11330, x.pr-2*x.se, col='green')
title("predicting 1 day")
# 1 week:
plot(meteo$T.outside,xlim=c(1,19970), type='l')
x.pr = as.numeric(predict(x, 10080)$pred)
x.se = as.numeric(predict(x, 10080)$se)
lines(9891:19970, x.pr, col='red')
lines(9891:19970, x.pr+2*x.se, col='green')
lines(9891:19970, x.pr-2*x.se, col='green')
title("predicting 1 week")
```

#### *Exercise 9* {.unnumbered}
::: {.callout-caution}
## Exercise 9

*Where does, for long-term forecasts, the predicted value converge to?*
:::

Now compare this with prediction using an AR(6) model for the residual with respect to the daily cycle:

```{r fig.width=5, fig.height=4}
# 1 week, including trend:
plot(meteo$T.outside,xlim=c(1,19970), type='l')
x.an = arima(an, c(6,0,0)) # model the anomaly by AR(6)
x.pr = as.numeric(predict(x.an, 10080)$pred)
x.se = as.numeric(predict(x.an, 10080)$se)
hours.all = c(meteo$hours, max(meteo$hours) + (1:10080)/60)
T.per = 18.2-4.9*sin(pi*(hours.all+1.6)/12)
lines(T.per, col = 'blue')
hours.pr = c(max(meteo$hours) + (1:10080)/60)
T.pr = 18.2-4.9*sin(pi*(hours.pr+1.6)/12)
lines(9891:19970, T.pr+x.pr, col='red')
lines(9891:19970, T.pr+x.pr+2*x.se, col='green')
lines(9891:19970, T.pr+x.pr-2*x.se, col='green')
title("predicting 1 week")
```

#### *Exercise 10* {.unnumbered}
::: {.callout-caution}
## Exercise 10

*Where does now, for long-term forecasts, the predicted value converge to? Explain the difference to the upper model.*
:::

#### *Exercise 11* {.unnumbered}
::: {.callout-caution}
## Exercise 11

*Fit a periodic trend and an AR(3) model to the humidity data. Plot predictions for one week.*
:::