---
title: "Spatio-temporal Modelling, Lab 6, YYYY-MM-DD"
editor: visual
---

# Spatial modelling: introductory matter II - Spatial Correlation



## Spatial correlation: the h-scatterplot

Before we jump into h-scatterplots of spatial data, let us first (re-)consider lagged scatterplots for temporal data.



#### *Exercise 1* {.unnumbered}
::: {.callout-caution}
## Exercise 1

*Create the lagged scatterplots for the T.outside variable in the meteo data set, one for lag 1 and one for lag 10. Compute and report the corresponding correlations. (To do the scatterplot, you simply need to plot each value of T.outside at time* $t$ against its counterpart at time $t-1$ or $t-10$, respectively)
:::

::: {.callout-tip}
## Solution


```{r,warning=F}
load("./data/meteo.RData")
h1 = meteo$T.outside[2:length(meteo$T.outside)]
h0 = meteo$T.outside[1:(length(meteo$T.outside) - 1)]
plot(h0, h1, type = "p")
cor(h0, h1)
title("h-scatterplot with lag=1")
```

```{r,warning=F}
load("./data/meteo.RData")
h1 = meteo$T.outside[11:length(meteo$T.outside)]
h0 = meteo$T.outside[1:(length(meteo$T.outside) - 10)]
plot(h0, h1, type = "p")
cor(h0, h1)
title("h-scatterplot with lag=10")

```

Although the correlation is very high in both cases, we can see that it is lower for lag=10. This is also reflected in the scatterplots that show much more scatter for lag=10
:::



Read the help for function hscat in package gstat.

We will use this function to compute $h$-scatterplots. An $h$-scatterplot plots data pairs $z(x_i), z(x_j)$ for selected pairs that have $|x_i - x_j|$ fall within a given interval. The intervals are given by the third argument, (breaks).

::: panel-tabset
## sp

```{r, warning=FALSE}
library(rgdal)
library(gstat)
library(sp)
meuse <- readOGR("./data/meuse.gpkg")
hscat(log(zinc) ~ 1, meuse, c(0, 30, 80, 100, 200, 300, 500, 1000))
```

## sf

```{r}
library(sf)
meuse.sf <- st_read("./data/meuse.gpkg")
hscat(log(zinc) ~ 1, meuse.sf, c(0, 30, 80, 100, 200, 300, 500, 1000))
```
:::



#### *Exercise 2* {.unnumbered}
::: {.callout-caution}
## Exercise 2

*Why is there no distance interval* $0-30$ in the plot?
:::

::: {.callout-tip}
## Solution

Because there are no points with distances between 0 and 30.
:::



#### *Exercise 3* {.unnumbered}
::: {.callout-caution}
## Exercise 3

*How many point pairs can you form from 155 observations?*
:::

::: {.callout-tip}
## Solution

You can get $n*(n-1)/2$ point pairs. In this case you get 11935 pairs from 155 observations.
:::



::: panel-tabset
## sp

```{r,warning=FALSE}
hscat(log(zinc) ~ 1, meuse, c(0, 80, 120, 250, 500, 1200))
```

## sf

```{r}
hscat(log(zinc) ~ 1, meuse.sf, c(0, 80, 120, 250, 500, 1200))
```
:::



#### *Exercise 4* {.unnumbered}
::: {.callout-caution}
## Exercise 4

*The correlation steadily drops with larger distance. Why?*
:::

::: {.callout-tip}
## Solution

The values are becoming less related with increasing distance. Statistically speaking: the correlation decreases with increasing distance. See [Toblers law](http://en.wikipedia.org/wiki/Tobler's_first_law_of_geography).
:::



::: panel-tabset
## sp

```{r,warning=F}
hscat(log(zinc) ~ 1, meuse, c(0, 120, 300, 600, 1200))
```

## sf

```{r}
hscat(log(zinc) ~ 1, meuse.sf, c(0, 120, 300, 600, 1200))
```
:::



#### *Exercise 5* {.unnumbered}
::: {.callout-caution}
## Exercise 5

*How would you interpret the negative correlation for the largest lag?*
::: 

::: {.callout-tip}
## Solution

The correllation in the distance class 600-1200 is only -0.056. This is no significant negative correlation. It only means that for large distances correlation converges to zero.
:::



## Spatial correlation: the variogram cloud

As you could see, the variogram cloud above was used to get $h$-scatterplots. The variogram cloud is basically the plot (or collection of plotted points) of

$0.5(Z(x_i)-Z(x_j))^2$ against $h = |x_i-x_j|$

in words: for each point pair $(x_i),z(x_j)$ half the squared difference in measured values against the spatial separation distance.

Plot a variogram cloud for the `log(zinc)` data:

::: panel-tabset
## sp

```{r,warning=F}
plot(variogram(log(zinc) ~ 1, meuse, cloud = TRUE))
```

## sf

```{r}
plot(variogram(log(zinc) ~ 1, meuse.sf, cloud = TRUE))
```
:::



#### *Exercise 6* {.unnumbered}
::: {.callout-caution}
## Exercise 6

*Why are both distance and semivariance non-negative?*
:::

::: {.callout-tip}
## Solution

Distance is per definition always positive. The semivariance $0.5(Z(x_i)-Z(x_j))^2$ uses sqaured differences, so it is always positive, too.
:::



#### *Exercise 7* {.unnumbered}
::: {.callout-caution}
## Exercise 7

*Why is there so much scatter in this plot?*
:::

::: {.callout-tip}
## Solution

For every point pair the distance is plotted against semivariance. This means there is one point in the cloud for every possible point pair of the data set.
:::



#### *Exercise 8* {.unnumbered}
::: {.callout-caution}
## Exercise 8

*How many points does the variogram cloud contain if the separation distance is not limited?*
:::

::: {.callout-tip}
## Solution

If the seperation distance is not limited, all points in the dataset are used to build pairs. This means $n*(n-1)/2$ point pairs, in this case 11935 (see 6.3).
:::



## Spatial correlation: the variogram

A variogram for the log-zinc data is computed, printed and plotted by

::: panel-tabset
## sp

```{r,warning=F}
v = variogram(log(zinc) ~ 1, meuse)
v
plot(v)
```

## sf

```{r}
v = variogram(log(zinc) ~ 1, meuse.sf)
v
plot(v)
```
:::

#### *Exercise 9* {.unnumbered}
::: {.callout-caution}
## Exercise 9

*How is the variogram computed from the variogram cloud?*

a)  semivariance is averaged for each distance
b)  distance is averaged for each semivariance
c)  distance is averaged for each semivariance class
d)  semivariance and distance are averaged for each distance class
e)  semivariance and distance are averaged for each semivariance class
:::

::: {.callout-tip}
## Solution

The correct answer is d).
:::



## Sample variogram and variogram model

To detect, or model spatial correlation, we average the variogram cloud values over distance intervals (bins):

::: panel-tabset
## sp

```{r,warning=F}
lzn.vgm = variogram(log(zinc) ~ 1, meuse)
lzn.vgm
plot(lzn.vgm)
```

## sf

```{r}
lzn.vgm = variogram(log(zinc) ~ 1, meuse.sf)
lzn.vgm
plot(lzn.vgm)
```
:::

For the maximum distance (cutoff) and bin width, sensible defaults are chosen as one-third the largest area diagonal and 15 bins, but we can modify this when needed by setting cutoff and width:

::: panel-tabset
## sp

```{r,warning=FALSE}
lzn.vgm = variogram(log(zinc) ~ 1, meuse, cutoff = 1000, width = 50)
lzn.vgm
plot(lzn.vgm)
```

## sf

```{r}
lzn.vgm = variogram(log(zinc) ~ 1, meuse.sf, cutoff = 1000, width = 50)
lzn.vgm
plot(lzn.vgm)
```
:::



#### *Exercise 10* {.unnumbered}
::: {.callout-caution}
## Exercise 10

*What is the problem if we set width to e.g. a value of 5?*

a)  we get too many semivariance estimates
b)  the semivariance values are too large
c)  the semivariance values are too small
d)  the variability of the semivariance values is too large
:::

::: {.callout-tip}
## Solution

The correct answer is d).
:::



#### *Exercise 11* {.unnumbered}
::: {.callout-caution}
## Exercise 11

*What is the problem if we set the cutoff value to e.g. a value of 500 and width to 50?*

a)  we cannot see up to which distance the data are spatially correlated
b)  the semivariance values are too large
c)  the semivariance values are too small
d)  we cannot see if data are spatially correlated
:::

::: {.callout-tip}
## Solution

The correct answer is d). There would not be enough points in the cloud to see if data are spatially correlated.
:::



For kriging, we need a suitable, statistically valid model fitted to the sample variogram. The reason for this is that kriging requires the specification of semivariances for any distance value in the spatial domain we work with. Simply connecting sample variogram points, as in

::: panel-tabset
## sp

```{r,warning=FALSE}
lzn.vgm = variogram(log(zinc) ~ 1, meuse, cutoff = 1000, width = 50)
plot(gamma ~ dist, lzn.vgm)
lines(c(0, lzn.vgm$dist), c(0, lzn.vgm$gamma))
```

## sf

```{r}
lzn.vgm = variogram(log(zinc) ~ 1, meuse.sf, cutoff = 1000, width = 50)
plot(gamma ~ dist, lzn.vgm)
lines(c(0, lzn.vgm$dist), c(0, lzn.vgm$gamma))
```
:::

will *not* result in a valid model. Instead, we will fit a valid parametric model. Some of the valid parametric models are shown by e.g.

```{r,warning=FALSE}
show.vgms()
```

These models may be combined, and the most commonly used are Exponential or Spherical models combined with a Nugget model.

We will try a spherical model:

::: panel-tabset
## sp

```{r,warning=FALSE}
lzn.vgm = variogram(log(zinc) ~ 1, meuse)
plot(lzn.vgm)

lzn.mod = vgm(0.6, "Sph", 1000, 0.05)
plot(lzn.vgm, lzn.mod)
```

## sf

```{r}
lzn.vgm = variogram(log(zinc) ~ 1, meuse.sf)
plot(lzn.vgm)

lzn.mod = vgm(0.6, "Sph", 1000, 0.05)
plot(lzn.vgm, lzn.mod)
```
:::



#### *Exercise 12* {.unnumbered}
::: {.callout-caution}
## Exercise 12

*The range, 1000, relates to*

a)  the semivariance value at distance (nearly) zero
b)  the semivariance value reached when the variogram stops increasing
c)  the distance value at which the variogram stops increasing
:::

::: {.callout-tip}
## Solution

The correct answer is c).
:::



#### *Exercise 13* {.unnumbered}
::: {.callout-caution}
## Exercise 13

*The sill, 0.6 + 0.05, relates to*

a)  the semivariance value at distance (nearly) zero
b)  the semivariance value reached when the variogram stops increasing
c)  the distance value at which the variogram stops increasing
:::

::: {.callout-tip}
## Solution

The correct answer is b).
:::



#### *Exercise 14* {.unnumbered}
::: {.callout-caution}
## Exercise 14

*The nugget, 0.05, relates to*

a)  the semivariance value at distance (nearly) zero
b)  the semivariance value reached when the variogram stops increasing
c)  the distance value at which the variogram stops increasing
:::

::: {.callout-tip}
## Solution

The correct answer is a).
:::



The variogram model (Sph for spherical) and the three parameters were chosen by eye. We can fit the three parameters also automatically:

::: panel-tabset
## sp

```{r,warning=F}
lzn.vgm = variogram(log(zinc) ~ 1, meuse)
lzn.mod = fit.variogram(lzn.vgm, vgm(0.6, "Sph", 1000, 0.05))
lzn.mod

plot(lzn.vgm, lzn.mod)
```

## sf

```{r}
lzn.vgm = variogram(log(zinc) ~ 1, meuse.sf)
lzn.mod = fit.variogram(lzn.vgm, vgm(0.6, "Sph", 1000, 0.05))
lzn.mod

plot(lzn.vgm, lzn.mod)
```
:::

but we need to have good "starting" values, e.g. the following will not yield a satisfactory fit:

```{r,warning=FALSE}
lzn.misfit = fit.variogram(lzn.vgm, vgm(0.6, "Sph", 10, 0.05))

plot(lzn.vgm, lzn.misfit)
```



#### *Exercise 15* {.unnumbered}
::: {.callout-caution}
## Exercise 15

*Compute the variogram for four different direction ranges (hint: look into argument alpha for function variogram) and plot it in a single plot. Is the zinc process anisotropic? Try to explain what you find in terms of what you know about the process.*
:::

::: {.callout-tip}
## Solution

::: panel-tabset
## sp

```{r,warning=F}
plot(variogram(log(zinc)~1, meuse, alpha=c(0,45,90,135)))  
```

## sf

```{r}
plot(variogram(log(zinc)~1, meuse.sf, alpha=c(0,45,90,135)))

```
:::

It seems range and sill depend on direction (see especially the 45 degree plot). The process is anisotropic. The polutants are propagated by the water of the river. Probably the flow direction has some influence.
:::