---
title: "Spatio-temporal Modelling, Lab 5, 2022-11-30"
editor: visual
---

# Spatial modelling: introductory matter


## meuse

The meuse data set is a data set collected from the Meuse floodplain. The heavy metals analyzed are copper, cadmium, lead and zinc concentrations, extracted from the topsoil of a part of the Meuse floodplain, the area around Meers and Maasband (Limburg, The Netherlands). The data were collected during fieldwork in 1990 [@meuse1; @meuse2].
Let´s have a look at the data. If not done yet, download the data set from learnweb. The code below assumes the data to sit in subdirectory `data` of your working directory. Adapt the path to wherever you have stored the data on your disk.

```{r}
library(gstat)
library(sp)
library(rgdal)
meuse <- readOGR("./data/meuse.gpkg")
summary(meuse)
objects()
```


## Models: the formula interface; linear regression

Most statisticians (and many earth scientists as well) like to analyse data through models: models reflect the hypotheses we entertain, and from fitting models, analysing model output, and analysing graphs of fitted values and residuals we can learn from the data whether a given hypothesis was reasonable or not.

Many models are variations of regression models. Here, we will illustrate an example of a simple linear regression model involving the heavy metal pollution data in `meuse`. We will entertain the hypothesis that zinc concentration, present in variable `zinc`, depends linearly on distance to the river, present in `dist`. We can express dependency of y on x by a formula, coded as `y~x`. Formulas can have multiple right-hand sides for multiple linear regression, as `y~x1+x2`; they also can contain expressions (transformations) of variables, as in `log(y)~sqrt(x)`. See `?formula` for a full description of the full functionality.

To calculate a linear regression model for the meuse data:

```{r}
zinc.lm = lm(zinc ~ dist, meuse)
class(zinc.lm)
```

```{r}
summary(zinc.lm)
```

Do the following yourself:

```{r}
par(mfrow=c(1,2)) #to get two plots in one row
plot(zinc.lm)
```

Besides the variety of plots you now obtain, there are many further custom options you can set that can help analysing these data. When you ask help by `?plot`, it does not provide very helpful information. To get help on the plot method that is called when plotting an object of class `lm`, remember that the function called is `plot.lm`. Read the help of `plot.lm` by `?plot.lm` . You can customize the plot call, e.g. by

```{r, warning=F}
par(mfrow=c(1,2)) 
plot(zinc.lm, add.smooth = FALSE)
```

In this call, the name of the second function argument is added because the argument panel is in a very rear position. Adding the `FALSE` as a positional argument, as in

```{r,warning=F,eval=F}
par(mfrow=c(1,2)) 
 plot(zinc.lm, , , , , , , FALSE)
```

will not work as it comes after the `...` argument, which may contain an arbitrary number of arguments that are passed to underlying plot routines. But the following two commands (search for the difference!) both work:

```{r,warning=F}
 par(mfrow=c(1,2)) 
plot(zinc.lm, caption = c("Residuen ~ Modellierte Werte", "QQ-Plot mit Normalverteilung", 
    " Standardisierte Residuen ~ Modellierte Werte", "Cook's Distanz", "Residuen vs Leverage", 
    "Cook's Distanz vs Leverage"))
```

```{r,warning=F}
par(mfrow=c(1,2)) 
plot(zinc.lm, , c("Residuen ~ Modellierte Werte", "QQ-Plot mit Normalverteilung", 
    " Standardisierte Residuen ~ Modellierte Werte", "Cook's Distanz", "Residuen vs Leverage", 
    "Cook's Distanz vs Leverage"))
```



#### *Exercise 1* {.unnumbered}
::: {.callout-caution}
## Exercise 1

*For this regression model, which of the following violated assumptions underlying linear regression are true (if any):*

a)  none
b)  residuals are heteroscedastic (i.e., their variance is not constant over the range of fitted values)
c)  residuals are not normally distributed (their distribution is for example non-symmetric)
d)  residuals are heteroscedastic and not normally distributed

*and explain how you can tell that this is the case.*
:::

::: {.callout-tip}
## Solution

The correct answer is d).In the third plot (standardised residuals \~ fitted values) you can see that the residuals are heteroscedastic, i.e. their variance is not constant. In the second plot (QQ-Plot), you can see that the residuals are not normally distributed.
:::

The example above is a spatial model only to the extent that the explanatory variable `dist` in this case contains some spatial information (distance to the river). We are not using the spatial information contained in the `SpatialPointsDataFrame` with respect to the location (coordinates) of the heavy metal measurements. In the following, we will look at actual spatial data analysis starting with how to handle spatial data in R.

## Spatial data in R (I): the sp package

The R package **sp** [@R-sp] provides classes and methods for spatial data; currently it can deal with points, grids, lines and polygons. It further provides interfaces to various GIS functions such as overlay, projection and reading and writing of external GIS formats.

### Points

Try:

```{r,warning=F}
class(meuse)
```

```{r, warning=F}
summary(meuse)
```

```{r, warning=F}
sp.theme(TRUE)
spplot(meuse, "zinc", key.space = "right")
```

The plot now obtained shows points, but without reference where on earth they lie. You could e.g. add a reference by showing the coordinate system:

```{r,warning=F}
spplot(meuse, "zinc", key.space = "right", scales = list(draw = TRUE))
```

but this tells little about *what* we see. As another option, you can add geographic reference by e.g. lines of rivers, or administrative boundaries. In our example, we prepared a polygon of the Meuse river boundaries, and will add it as reference:

```{r,warning=F}
meuse.riv <- readOGR("./data/meuse_river.gpkg")
```

```{r,warning=F}
meuse.lt = list("sp.polygons", meuse.riv, fill = "grey")
spplot(meuse, "zinc", key.space = "right", sp.layout = meuse.lt)
```

The `sp.layout` argument contains map elements to be added to an `spplot` plot.

Note that the points in the plot partly fall in the river; this may be attributed to one or more of the following reasons: (i) the river coordinates are not correct, (ii) the point coordinates are not correct, (iii) points were taken on the river bank when the water was low, whereas the river boundary indicates the high water extent of the river.

### Grids

Try:

```{r,warning=F}
library(rgdal)
meuse.grid <- as(readGDAL("./data/meuse_grid.tif"),"SpatialPixelsDataFrame")
class(meuse.grid)
```

As you can see, the GeoTiff has been loaded as a `SpatialPixelsDataFrame`. The GeoTiff contains 5 bands: bands 1 and 2 simply indicate, whether a pixel belongs to "part.a" or "part.b" of the study area ("yes" or "no", i.e., 1 or 0 in the corresponding band). This is an arbitrary division of the area in two parts. Band 3 shows the distance to the river Meuse, band 4 has the soil type, and band 5 indicates the number of times a pixel has been flooded before. To represent this in our `SpatialPixelsDataFrame`, we can assign corresponding names to the data attributes. We will also change the type of the attributes containing categorical data to `factor`.

```{r}
names(meuse.grid) <- c("part.a","part.b","dist","soil","ffreq")
meuse.grid$soil <- as.factor(meuse.grid$soil)
meuse.grid$ffreq <- as.factor(meuse.grid$ffreq)
```

```{r,warning=F}
summary(meuse.grid)
```
Let´s try to plot the data:
```{r,warning=F}
meuse.lt = list(riv = list("sp.polygons", meuse.riv, fill = "grey"), pts = list("sp.points", 
    meuse, pch = 3, cex = 0.5, col = "black"))
spplot(meuse.grid, sp.layout = meuse.lt, key.space = "right")
```

For explanation on the `sp.layout` argument, read `?spplot`; much of the codes in them (`pch, cex, col`) are documented in `?par`.

Note that `spplot` plots points on top of the grid, and the grid cells on top of the polygon with the river. (When printing to pdf, transparency options can be used to combine both.)

## Import/export: rgdal

R package `rgdal` provides several drivers for import/export of spatial data. We have used it above to read the vector and raster data `meuse`, `meuse.riv` and `meuse.grid`. An overview of available drivers for vector data (points, lines, polygons) can be obtained by

```{r,warning=F}
ogrDrivers()
```

and for grid data by

```{r,warning=F}
gdalDrivers()
```

The full set of theoretically available drivers with documentation are given [here](http://www.gdal.org/formats_list.html) and [here](http://www.gdal.org/ogr_formats.html).

Export the data set to KML

```{r,warning=F,eval=F}
 library(rgdal)
 data(meuse)
 coordinates(meuse) = ~x+y
 proj4string(meuse) = CRS("+init=epsg:28992")
 proj4string(meuse)
 meuse.ll = spTransform(meuse, CRS("+proj=longlat"))
 writeOGR(meuse.ll, "meuse.kml", "meuse.kml", driver="KML")
```

and import it in google earth.

## Spatial data in R (II): the packages sf and stars

The package **sf** [@R-sf], which first appeared on CRAN in 2016, is intended to replace the vector parts of **sp** (among others). It was developed "to move spatial data analysis in R closer to standards-based approaches seen in the industry and open source projects, to build upon more modern versions of the open source geospatial software stack [...], and to allow for integration of R spatial software with the tidyverse [...]" [@sdsr, chap. 7.1].  
The package **stars** [@R-stars], which first came to CRAN in 2018, was developed for analyzing raster (and vector) data cubes. It is meant to extend current packages for raster data with functionality to represent dynamic raster stacks (e.g. time series) and to scale computations to data beyond local disc sizes [@sdsr, chap. 7.4].

Both packages are developed to work closely together and can be used for all spatial analyses throughout our exercises. In the following, we show how to repeat the examples from above for handling points and grids with **sf** and **stars**.

For an introduction to **sf** and **stars**, see

### Points

```{r}
library(sf)
meuse.sf <- st_read("./data/meuse.gpkg")

```
```{r}
class(meuse.sf)
```

```{r}
summary(meuse.sf)
```

```{r}
plot(meuse.sf)
```

```{r}
plot(meuse.sf["zinc"],axes=T)
```

Again, we can add the meuse river as refeference.

```{r}
meuse.riv.sf <- st_read("./data/meuse_river.gpkg")
plot(meuse.sf["zinc"],axes=T,reset=F)
plot(meuse.riv.sf,col='grey',add=T)
```

### Grids

Try: 

```{r}
library(stars)
meuse.grid.stars <- read_stars("./data/meuse_grid.tif")
meuse.grid.stars
```
The bands of our data set are automatically interpreted as another dimension of the data (here you can recognize the characteristic of **stars** to naturally support multidimensional raster data). However, in this case our bands do not represent another dimension (like for example having the same measurements at different times in the different bands), so instead we want them to be interpreted as 5 different attributes of a 2-dimensional array. This can be done using the function `split`:

```{r}
meuse.grid.stars <- split(meuse.grid.stars, "band")
meuse.grid.stars

```

Similar to above, we can now assign the names to the attributes and change the type of the attributes representing categorical variables to `factor`.

```{r}
names(meuse.grid.stars) <- c("part.a","part.b","dist","soil","ffreq")
meuse.grid.stars$soil <- as.factor(meuse.grid.stars$soil)
meuse.grid.stars$ffreq <- as.factor(meuse.grid.stars$ffreq)
meuse.grid.stars
```

Let´s try a simple plot before looking more into visualization of `sf` and `stars`objects.

```{r}
plot(meuse.grid.stars[3], col=heat.colors(10))
```

## Visualizing spatial data with the tmap package

In the previous section, we have used the function `spplot()` to create plots from data handled by the package **sp**, such as `SpatialPointsDataFrame` or `SpatialPixelsDataFrame`. While there are base `plot` methods for `sf` and `stars` objects as well (we have used some of them in the examples above, more examples can be found in the lecture material), we will use the package **tmap** [@R-tmap]  for plotting objects of those classes throughout this exercise. **tmap** allows easy handling of `sf` and `stars` objects combined in one plot and offers a lot of utility for creating and styling maps and other visualizations of spatial data. The syntax for plotting is very simple and straightforward: the basic building blog of a plot is a `tm_shape` element, which defines the spatial data object (raster or vector) to be added to the map plus one or more layer elements, e.g., `tm_raster` or `tm_polygons`, which define how the object is to be plotted (and which aspects of it). Spatial data objects can be, e.g., an `sf` or `stars` object, but also objects from the packages **sp** or **raster**. All these elements are combined via the + operator. The example below creates a plot with one layer showing the "dist" attribute of `meuse.grid.stars` together with the polygon of the Meuse river. 

```{r}
library(tmap)
tm_shape(meuse.grid.stars)+tm_raster("dist",n=10)+tm_shape(meuse.riv.sf)+tm_polygons() #plots the `dist` attribute of the spatial data object `meuse.grid.stars` and the polygon of the Meuse river contained in spatial data object `meuse.riv.sf` 
```

The functions for adding the layers like `tm_raster()` or `tm_polygons()` allow for several arguments further specifying how the layers should be visualized. In the code above we only specified that the raster layer should show the "dist" attribute with a preferred number of 10 classes. Many other arguments can be set as well (see below for an example that adds the point data of the "meuse.sf" data set with symbols of a specific shape, size and color).

```{r}
bpy.palette <- bpy.colors() #define color palette for the following plots equal to the one used above by spplot()
tm_shape(meuse.grid.stars) + 
  tm_raster("dist",n=10, palette = bpy.palette) + 
  tm_shape(meuse.riv.sf) + 
  tm_polygons() + 
  tm_shape(meuse.sf) + 
  tm_symbols(shape = 3,size = 0.2,col="green")
```

There are also other tmap elements, e.g., `tm_layout()`, which further specifies the overall map layout, or `tm_facets()`, which can be used to create small multiples in a map.

```{r}
tm_shape(meuse.grid.stars) + 
  tm_raster(c("dist","soil")) + 
  tm_shape(meuse.riv.sf)+ tm_polygons() + 
  tm_shape(meuse) + tm_symbols(shape = 3,size = 0.2,col="black") + 
  tm_facets() + 
  tm_layout(legend.outside.size = 0.15, main.title = "distance to river and soil type in one plot", panel.labels = c("distance to river","soil type"))
```


**tmap** can be used not only for plotting static maps, but also for visualizing spatial data in an interactive viewer. Changing the mode to "view" will cause all visualizations to be shown in an interactive leaflet map including selectable background layers as well as zoom and pan functionality. The following code shows an example (note that after creating the interactive map, we switch back to the mode "plot" for creating static maps in the following sections).

```{r}
tmap_mode("view")
tm_shape(meuse.grid.stars) + 
  tm_raster("dist",n=10) + 
  tm_shape(meuse.riv)+tm_polygons()
tmap_mode("plot")
```

For more information on the use of **tmap** please refer to the [get started!](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html#interactive-maps) vignette.


## sp vs. sf & stars

Whether you prefer to use **sp** or **sf** and **stars**, is up to you. As mentioned above **sf** and **stars** are intended to replace **sp** in the long term. For now, we will always present the code for both alternatives when analyzing spatial data. You can switch between them by selecting the corresponding tabs in the code sections. You will notice that for the most part there are only minor (if any) differences between both alternatives. Most of the functions we will be using can handle objects from **sp**, **sf** and **stars** identically. Main differences occur when plotting the data (`spplot` for **sp**, functions from **tmap** for **sf** & **stars**).  

# Geostatistics

The following practical exercises cover the use of geostatistical applications in environmental research. They give an introduction to exploratory data analysis, sample variogram calculation and variogram modelling and several spatial interpolation techniques. Interpolation methods described and applied are inverse distance interpolation, trend surface fitting, thiessen polygons, ordinary (block) kriging, stratified (block) kriging and indicator (block) kriging.

The **gstat** R package [@R-gstat] is used for all geostatistical calculations. See @gstatcg for more info.


## Simple interpolation algorithms

### Trend surface analysis

Trend surface interpolation is multiple linear regression with polynomials of coordinates as independent, predictor variables. Although we could use `lm` to fit these models and predict them, it would require scaling of coordinates prior to prediction in most cases because higher powers of large coordinate values result in large numerical errors. The `krige` function in library `gstat` scales coordinates and computes trend surface up to order 3 (argument `degree`), using scaled coordinates.

```{r,warning=F}


```

::: panel-tabset

## sp

```{r}

meuse.grid$tr1 = krige(log(zinc) ~ 1, meuse, meuse.grid, degree = 1)$var1.pred
meuse.grid$tr2 = krige(log(zinc) ~ 1, meuse, meuse.grid, degree = 2)$var1.pred
meuse.grid$tr3 = krige(log(zinc) ~ 1, meuse, meuse.grid, degree = 3)$var1.pred
spplot(meuse.grid, c("tr1", "tr2", "tr3"), sp.layout = meuse.lt, main = "log(zinc), trend surface interpolation")

```

## sf/stars

```{r}
meuse.grid.stars$tr1 = krige(log(zinc) ~ 1, meuse.sf, meuse.grid.stars, degree = 1)$var1.pred
meuse.grid.stars$tr2 = krige(log(zinc) ~ 1, meuse.sf, meuse.grid.stars, degree = 2)$var1.pred
meuse.grid.stars$tr3 = krige(log(zinc) ~ 1, meuse.sf, meuse.grid.stars, degree = 3)$var1.pred
tm_shape(meuse.grid.stars) + 
  tm_raster(c("tr1","tr2","tr3"),n=20,title = "prediction", palette = bpy.palette) + 
  tm_shape(meuse.riv.sf)+tm_polygons() + 
  tm_shape(meuse)+tm_symbols(shape = 3,size = 0.2,col="black") + 
  tm_facets(free.scales = FALSE) + 
  tm_layout(legend.outside.size = 0.15, main.title = "log(zinc), trend surface interpolation", panel.labels = c("tr1","tr2","tr3"))


```

:::

::: {.callout-note}
Function `surf.ls` from library `spatial` fits trends up to order 6 with scaled coordinates.
:::


Here is another example of the use of the `...` argument in function definitions. As you can see, `?krige` does not provide help on argument `degree`. However, it tells that any remaining arguments (`...`) are passed to function `gstat`, so look for argument `degree` on `?gstat` to see information on this argument.

In the example above, we have fitted trends up to order 3. The higher the order, the more exactly the surface can be fitted to the data. However...

#### *Exercise 2* {.unnumbered}
::: {.callout-caution}
## Exercise 2

*The danger of extreme extrapolation values is likely to occur*

a)  for low trend orders
b)  for high trend orders
c)  for none of these models

*Briefly explain why this is the case.*
:::

::: {.callout-tip}
## Solution

The correct answer is b). If you plot polynomials in one dimension, you can easily see that function values are more extreme for higher trend orders.

```{r,warning=F}
x = seq(-2, 2, 0.1)
plot(x, x, type = "l", ylim = c(-2, 2), col = "green")  # 1. Ordnung (Gerade)
abline(v = 0, h = 0)
lines(x, x^2, col = "red")  # 2. Ordnung (Parabel)
lines(x, x^3, col = "blue")  # 3. Ordnung
```
:::



Local trends, i.e. trends in a local neighbourhood may also be fitted:

:::panel-tabset

## sp
```{r}
m = krige(log(zinc) ~ 1, meuse, meuse.grid, degree=1, nmax = 10)
spplot(m, "var1.pred", sp.layout = meuse.lt, main = "local 1st order trend")
```

## sf/stars

```{r}
m = krige(log(zinc) ~ 1, meuse.sf, meuse.grid.stars, degree=1, nmax = 10)
tm_shape(m) + 
  tm_raster("var1.pred", n=20, title = "prediction", palette = bpy.palette) + 
  tm_shape(meuse.riv.sf)+tm_polygons() + 
  tm_shape(meuse)+tm_symbols(shape = 3,size = 0.2,col="black") + 
  tm_layout(main.title = "local 1st order trend", legend.outside = TRUE)

```

:::

### Inverse distance interpolation

Inverse distance interpolation calculates a weighted average of points in the (by default global) neighbourhood, using weights inverse proportional to the distance of data locations to the prediction location raised to the power $p$. This power is by default 2, but can of course be adapted:

:::panel-tabset

## sp

```{r,warning=F}

lzn.tp = idw(log(zinc) ~ 1, meuse, meuse.grid)

spplot(lzn.tp, "var1.pred", sp.layout = meuse.lt, main = "log(zinc), inverse distance interpolation")

meuse.grid$idp0.5 = idw(log(zinc) ~ 1, meuse, meuse.grid, idp = 0.5)$var1.pred

meuse.grid$idp02 = idw(log(zinc) ~ 1, meuse, meuse.grid, idp = 2)$var1.pred

meuse.grid$idp05 = idw(log(zinc) ~ 1, meuse, meuse.grid, idp = 5)$var1.pred

meuse.grid$idp10 = idw(log(zinc) ~ 1, meuse, meuse.grid, idp = 10)$var1.pred

spplot(meuse.grid, c("idp0.5", "idp02", "idp05", "idp10"), sp.layout = meuse.lt, 
    main = "log(zinc), inverse distance interpolation")
```

## sf/stars

```{r}
lzn.tp = idw(log(zinc) ~ 1, meuse.sf, meuse.grid.stars)

tm_shape(m)+tm_raster("var1.pred", palette = bpy.palette, n=10) + 
  tm_shape(meuse.riv.sf) + 
  tm_polygons() + 
  tm_shape(meuse) +
  tm_symbols(shape = 3,size = 0.2,col="black") +
  tm_layout(main.title = "log(zinc), inverse distance interpolation", legend.outside = TRUE, title = "prediction" )

meuse.grid.stars$idp0.5 = idw(log(zinc) ~ 1, meuse, meuse.grid.stars, idp = 0.5)$var1.pred

meuse.grid.stars$idp02 = idw(log(zinc) ~ 1, meuse, meuse.grid.stars, idp = 2)$var1.pred

meuse.grid.stars$idp05 = idw(log(zinc) ~ 1, meuse, meuse.grid.stars, idp = 5)$var1.pred

meuse.grid.stars$idp10 = idw(log(zinc) ~ 1, meuse, meuse.grid.stars, idp = 10)$var1.pred

tm_shape(meuse.grid.stars) +
  tm_raster(c("idp0.5","idp02","idp05","idp10"),n=10, title = "prediction", palette = bpy.palette) +
  tm_shape(meuse.riv.sf) + 
  tm_polygons() + 
  tm_shape(meuse) +
  tm_symbols(shape = 3,size = 0.2,col="black") +
  tm_facets(nrow=1,free.scales = FALSE) +
  tm_layout(main.title = "log(zinc), inverse distance interpolation",legend.outside.size = 0.15,panel.labels = c("idp0.5","idp02","idp05","idp10"))
```


:::


#### *Exercise 3* {.unnumbered}
::: {.callout-caution}
## Exercise 3

*With a larger inverse distance power (`idp`), the weight of the nearest data point*

a)  becomes larger
b)  becomes smaller
c)  does not change

*explain in words why this is the case.*
:::

::: {.callout-tip}
## Solution

The correct answer is a). With a higher inverse distance power the weight of a data point decreases more quickly with increasing distance. $$\lambda_i = \frac{|s_i-s_0|^{-p}}{\sum_{i=1}^n |s_i-s_0|^{-p}}$$

:::



### Nearest Neighbor / Tiessen (or Voronoi) Polygons

Nearest neighbor interpolation on a regular grid (creating Thiessen Polygons) is simply obtained by using e.g. inverse distance interpolation and a neighbourhood size of one (`nmax = 1`). :

:::panel-tabset

## sp

```{r,warning=F}
lzn.tp = krige(log(zinc) ~ 1, meuse, meuse.grid, nmax = 1)
spplot(lzn.tp, "var1.pred", sp.layout = meuse.lt, main = "Thiessen (or Voronoi) polygon interpolation of log(zinc)")
```

## sf/stars

```{r}
lzn.tp = krige(log(zinc) ~ 1, meuse.sf, meuse.grid.stars, nmax = 1)
tm_shape(lzn.tp) + 
  tm_raster("var1.pred",n=10,title="prediction", palette = bpy.palette) +
  tm_shape(meuse.riv.sf)+tm_polygons() +
  tm_shape(meuse) +
  tm_symbols(shape = 3,size = 0.2,col="black")
```

:::


# Spatial prediction with multiple linear regression

In the data set `meuse` we have a large number of variables present that can be used to form predictive regression models for either of the heavy metal variables. If we want to use such a regression model for spatial prediction, we need the variables as a full spatial coverage, e.g. in the form of a regular grid covering the study area, as well.

Only few variables are available for this in `meuse.grid`:

:::panel-tabset

## sp

```{r, warning=F}
meuse.grid <- as(readGDAL("./data/meuse_grid.tif"),"SpatialPixelsDataFrame") #reload original data set
names(meuse.grid) <- c("part.a","part.b","dist","soil","ffreq")
meuse.grid$soil <- as.factor(meuse.grid$soil)
meuse.grid$ffreq <- as.factor(meuse.grid$ffreq)
names(meuse.grid)
```
## sf/stars
```{r}
meuse.grid.stars <- read_stars("./data/meuse_grid.tif") #reload original data set
meuse.grid.stars <- split(meuse.grid.stars, "band")
names(meuse.grid.stars) <- c("part.a","part.b","dist","soil","ffreq")
meuse.grid.stars$soil <- as.factor(meuse.grid.stars$soil)
meuse.grid.stars$ffreq <- as.factor(meuse.grid.stars$ffreq)
names(meuse.grid.stars)
```

:::


Now try:

:::panel-tabset

## sp

```{r}

lzn.lm = lm(log(zinc) ~ sqrt(dist), meuse)
summary(lzn.lm)


par(mfrow=c(1,2)) 
plot(lzn.lm)

plot(log(zinc) ~ sqrt(dist), meuse)
abline(lzn.lm)


lzn.pr = predict(lm(log(zinc) ~ sqrt(dist), meuse), meuse.grid, se.fit = TRUE)
meuse.grid$lzn.fit = lzn.pr$fit
spplot(meuse.grid, "lzn.fit")

meuse.grid$se.fit = lzn.pr$se.fit
spplot(meuse.grid, "se.fit")
```

## sf/stars

```{r}

lzn.lm = lm(log(zinc) ~ sqrt(dist), meuse.sf)
summary(lzn.lm)


par(mfrow=c(1,2)) 
plot(lzn.lm)

plot(log(zinc) ~ sqrt(dist), meuse.sf)
abline(lzn.lm)


lzn.pr = predict(lm(log(zinc) ~ sqrt(dist), meuse.sf), meuse.grid.stars, se.fit = TRUE)
meuse.grid.stars$lzn.fit = lzn.pr$fit


tm_shape(meuse.grid.stars)+tm_raster("lzn.fit",n=10, palette = bpy.palette)
meuse.grid.stars$se.fit = lzn.pr$se.fit
tm_shape(meuse.grid.stars)+tm_raster("se.fit", n=10, palette = bpy.palette)

```

:::



#### *Exercise 4* {.unnumbered}
::: {.callout-caution}
## Exercise 4

*Why are prediction standard errors on the log-scale largest for small and large distance to the river?*

a)  Because variability is smallest for intermediate distance to river
b)  because the uncertainty in the regression slope is most prevalent when the regressor *distance* takes extreme values
c)  because we have too few degrees of freedom for regression (i.e., too few observations) overall
d)  because we have too little data close to the river and far away from the river
:::

::: {.callout-tip}
## Solution

The correct answer is b). The uncertainty introduced by the slope of the regression model is highest for low and high distance values.
:::



We can use the prediction errors provided by `lm` to calculate regression prediction intervals:

:::panel-tabset

## sp

```{r,warning=F}
plot(log(zinc) ~ sqrt(dist), meuse)
x = 0:100/100
lzn.lm = lm(log(zinc) ~ sqrt(dist), meuse)
pr = predict(lzn.lm, data.frame(dist = x), interval = "prediction")
abline(lzn.lm)
lines(sqrt(x), pr[, 2], lty = "dashed", col = "red")
lines(sqrt(x), pr[, 3], lty = "dashed", col = "red")
pr = predict(lzn.lm, data.frame(dist = x), interval = "confidence")
lines(sqrt(x), pr[, 2], lty = "dashed", col = "blue")
lines(sqrt(x), pr[, 3], lty = "dashed", col = "blue")
title("95% confidence intervals (blue) and prediction intervals (red)")
```

## sf/stars

```{r}
plot(log(zinc) ~ sqrt(dist), meuse.sf)
x = 0:100/100
lzn.lm = lm(log(zinc) ~ sqrt(dist), meuse.sf)
pr = predict(lzn.lm, data.frame(dist = x), interval = "prediction")
abline(lzn.lm)
lines(sqrt(x), pr[, 2], lty = "dashed", col = "red")
lines(sqrt(x), pr[, 3], lty = "dashed", col = "red")
pr = predict(lzn.lm, data.frame(dist = x), interval = "confidence")
lines(sqrt(x), pr[, 2], lty = "dashed", col = "blue")
lines(sqrt(x), pr[, 3], lty = "dashed", col = "blue")
title("95% confidence intervals (blue) and prediction intervals (red)")
```
:::


Next, we can make maps of the 95% prediction intervals:


:::panel-tabset

## sp
```{r,warning=F}
pr = predict(lzn.lm, as.data.frame(meuse.grid), interval = "prediction")
meuse.grid$fit = pr[, 1]
meuse.grid$upper = pr[, 3]
meuse.grid$lower = pr[, 2]
spplot(meuse.grid, c("fit", "lower", "upper"), sp.layout = meuse.lt, layout = c(3, 
    1))
```

## sf/stars

```{r}
pr = predict(lzn.lm, as.data.frame(meuse.grid.stars), interval = "prediction")
meuse.grid.stars$fit = pr[, 1]
meuse.grid.stars$upper = pr[, 3]
meuse.grid.stars$lower = pr[, 2]
tm_shape(meuse.grid.stars) + 
  tm_raster(c("fit","lower","upper"),n=20,title = "prediction", palette = bpy.palette) + 
  tm_shape(meuse.riv.sf)+tm_polygons() + 
  tm_shape(meuse)+tm_symbols(shape = 3,size = 0.2,col="black") +
  tm_facets(ncol = 3,nrow=1,free.scales = FALSE) + 
  tm_layout(legend.outside.size = 0.15, panel.labels = c("tr1","tr2","tr3"))
```

:::


Note that all these prediction intervals refer to prediction of single observations; confidence intervals for mean values (i.e., the regression line alone) are obtained by specifying interval = `"confidence"`.



#### *Exercise 5* {.unnumbered}
::: {.callout-caution}
## Exercise 5

*Compared to prediction intervals, the intervals obtained by interval = `"confidence"` are*

a)  wider
b)  narrower
c)  of approximately equal width
:::

::: {.callout-tip}
## Solution

The correct answer is b). The confidence intervals for **single observations** (prediction intervals) are wider than intervals obtained by `interval = confidence"` because the latter are for the regression line, i.e. confidence intervals of the **mean** for a certain distance.
:::



Another regression model is obtained by relating zinc to flood frequency, try

:::panel-tabset

## sp

```{r,warning=F}
plot(log(zinc) ~ ffreq, meuse)
meuse$ffreq <- as.factor(meuse$ffreq)
plot(log(zinc) ~ ffreq, meuse)
```


## sf/stars
```{r}
plot(log(zinc) ~ ffreq, meuse.sf)
meuse.sf$ffreq <- as.factor(meuse.sf$ffreq)
plot(log(zinc) ~ ffreq, meuse.sf)

```

:::




#### *Exercise 6* {.unnumbered}
::: {.callout-caution}
## Exercise 6

*Why is a box-and whisker plot drawn for the second plot, and not a scatter plot?*

a)  `ffreq` has been turned into a categorical variable with numeric levels
b)  `ffreq` has been turned into a variable with only three values, 1, 2 and 3
c)  flood frequency is always expressed by boxes
:::

::: {.callout-tip}
## Solution

The correct answer is a). Flood frequency in our data set has numbers as labels, but can actually be seen as a categorical variable with the number of floodings being the categories. By using `as.factor()` we explicitly turn this variable into a factor, R´s representation of categorical variables. When plotting such a variable, a more meaningful representation (box plots) is automatically used.
:::



#### *Exercise 7* {.unnumbered}
::: {.callout-caution}
## Exercise 7

*What do the boxes in a box-and whisker plot refer to?*

a)  the mean plus and minus one standard deviation of the group
b)  the 95 percent confidence interval for the mean
c)  the inter-quartile range (from 25-th to 75-th percentiles)
:::

::: {.callout-tip}
## Solution

The correct answer is c).
:::



#### *Exercise 8* {.unnumbered}
::: {.callout-caution}
## Exercise 8

*What is needed for flood frequency to use it for mapping zinc?*

a)  it needs to be the number of times a grid cell floods
b)  it needs to be known at each prediction location (i.e., grid cells in `meuse.grid`)
c)  it needs to be independent of distance to the river
:::

::: {.callout-tip}
## Solution

The correct answer is b).
:::



Try the model in prediction mode:

:::panel-tabset

## sp

```{r,warning=F}
lzn.lm2 = lm(log(zinc) ~ ffreq, meuse)
summary(lzn.lm2)

pr = predict(lzn.lm2, meuse.grid, interval = "prediction")
meuse.grid$fit = pr[, 1]
meuse.grid$upper = pr[, 3]
meuse.grid$lower = pr[, 2]
spplot(meuse.grid, c("fit", "lower", "upper"), sp.layout = meuse.lt, layout = c(3,1))
```

## sf/stars
```{r}
lzn.lm2 = lm(log(zinc) ~ ffreq, meuse.sf)
summary(lzn.lm2)

pr = predict(lzn.lm2, meuse.grid.stars, interval = "prediction")
meuse.grid.stars$fit = pr[, 1]
meuse.grid.stars$upper = pr[, 3]
meuse.grid.stars$lower = pr[, 2]
tm_shape(meuse.grid.stars) + 
  tm_raster(c("fit", "lower", "upper"), palette = bpy.palette, title = "", n=20) + 
  tm_shape(meuse.riv)+tm_polygons() + 
  tm_shape(meuse.sf) + 
  tm_symbols(size = 0.2,shape = 3,col = "black") + 
  tm_facets(free.scales = FALSE) +
  tm_layout(legend.outside.size = 0.15, panel.labels = c("fit","lower","upper"))
```

:::



#### *Exercise 9* {.unnumbered}
::: {.callout-caution}
## Exercise 9

*Why is the pattern obtained not smooth?*
:::

::: {.callout-tip}
## Solution

ffreq is a categorical variable so there are leaps in the map where the flood frequency changes.
:::


